% Copyright (c) 2016, The Science and Technology Facilities Council (STFC)
% All rights reserved.
% Error and warning messages -- same for C and Fortran

Possible values are:
\begin{description}
\item{} {\tt -1} Maximum number of iterations reached without convergence.
\item{} {\tt -2} Error from evaluating a function/Jacobian/Hessian.
\item{} {\tt -3} Unsupported choice of model.
\item{} {\tt -4} Error return from an external routine.
\item{} {\tt -5} Unsupported choice of method.
\item{} {\tt -6} Allocation error.
\item{} {\tt -7} Maximum number of reductions of the trust radius reached.
\item{} {\tt -8} No progress being made in the solution.
\item{} {\tt -9} \texttt{n}$>$\texttt{m}.
\item{} {\tt -10} Unsupported trust region update strategy.
\item{} {\tt -11} Unable to valid step when solving trust region subproblem.
\item{} {\tt -12} Unsupported scaling method.
\item{} {\tt -13} Error accessing pre-allocated workspace.
\item{} {\tt -101} Unsupported model in dogleg (\texttt{nlls\_method = 1}).
\item{} {\tt -201}  All eigenvalues are imaginary (\texttt{nlls\_method=2}).
\item{} {\tt -202} Matrix with odd number of columns sent to \texttt{max\_eig} subroutine (\texttt{nlls\_method=2}).
\item{} {\tt -301} {\tt nlls\_options\ct more\_sorensen\_max\_its} is exceeded in \texttt{more\_sorensen} subroutine (\texttt{nlls\_method=3}).
\item{} {\tt -302} Too many shifts taken in \texttt{more\_sorensen} subroutine (\texttt{nlls\_method=3}).
\item{} {\tt -303} No progress being made in \texttt{more\_sorensen} subroutine (\texttt{nlls\_method=3}).
\item{} {\tt -401} {\tt nlls\_options\ct model = 4} selected, but {\tt nlls\_options\ct exact\_second\_derivatives} is set to {\tt false}.
\item{} {\tt -501} {\tt nlls\_options\ct type\_of\_method = 2} selected, but {\tt nlls\_options\ct type\_of\_method} is not equal to {\tt 4}.
\end{description}
