% Copyright (c) 2016, The Science and Technology Facilities Council (STFC)
% All rights reserved.
\fortranonly{
The derived data type {\tt NLLS\_inform} is used to hold information from the execution of {\tt nlls\_solve} and {\tt nlls\_iterate}.
}
\conly{
The structure of type {\tt struct ral\_nlls\_inform} is used to hold information from the execution of {\tt ral\_nlls}.
}

The components are:
\begin{description}
\scalarinteger{status}
that gives the exit status of the subroutine.  See Section~\ref{hslerrors} for details.
\scalarcharacter{error\_message}{80} that holds the error message corresponding to the exit status.
\scalarinteger{alloc\_status} that gives the status of the last attempted allocation/deallocation.
\scalarcharacter{bad\_alloc}{80} that holds the name of the array that was being allocated when an error was flagged.
\scalarinteger{iter} that gives the total number of iterations performed.
\scalarinteger{f\_eval} that gives the total number of evaluations of the objective function.
\scalarinteger{g\_eval} that gives the total number of evaluations of the gradient of the objective function.
\scalarinteger{h\_eval} that gives the total number of evaluations of the Hessian of the objective function.
\scalarinteger{convergence\_normf} that tells us if the test on the size of \(\vr\) is satisfied.
\scalarinteger{convergence\_normg} that tells us if the test on the size of the gradient is satisfied.
\scalarinteger{convergence\_norms} that tells us if the test on the step length is satisfied.
\ronearrayreal{resvec}{iter+1} that, if {\tt nlls\_options\ct output\_progress\_vectors=true}, holds the vector of residuals.
\ronearrayreal{gradvec}{iter+1} that, if {\tt nlls\_options\ct output\_progress\_vectors=true}, holds the vector of gradients.
\scalarreal{obj} that holds the value of the objective function at the best estimate of the solution determined by the algorithm.
\scalarreal{norm\_g} that holds the gradient of the objective function at the best estimate of the solution determined by the package.
\scalarreal{scaled\_g} that holds the gradient of the objective function at the best estimate of the solution determined by the package.
\scalarinteger{external\_return} that gives the error code that was returned by a call to an external routine.
\scalarcharacter{external\_name}{80} that holds the name of the external code that flagged an error.
\scalarreal{step} holds the size of the last step taken.
\end{description}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "nlls_fortran"
%%% End:
