% the options derived data type

\fortranonly{
The derived data type {\tt NLLS\_options}
is used to hold controlling data. The components  are automatically
given default values in the definition of the type.
}
\conly{
The structure of type {\tt struct ral\_nlls\_options} is used to hold
controlling data. The components must be initialised through a call to
\texttt{ral\_nlls\_default\_options()}.
}

\vspace{2mm}

\noindent {\bf Components that control printing}
\begin{description}

\scalarinteger{out}
with default value {\tt 6} that is used as the \conly{Fortran }unit number for general messages. If it is negative, these messages will be suppressed.

\scalarinteger{print\_level} 
with default value {\tt 0} that controls the level of output required.
\begin{description}
\item{\tt $\leq$ 0} No informational output will occur.
\item{\tt = 1} Gives a one-line summary for each iteration.
\item{\tt = 2} As 1, plus gives a summary of the inner iteration for each iteration.
\item{\tt > 2} As 2, and gives increasingly verbose (debugging) output.
\end{description}
The default is {\tt print\_level} $=$ 0.
\end{description}

\scalarlogical{print_options}
with default value {\tt false}, determines whether to print a list of all options and their values at the beggining of the solve.

\scalarinteger{print_header}
with default value {\tt 30}, it will print the column header every 30 iterations when {\tt print\_level} is greater than 1.

\noindent {\bf Components that control the main iteration}.

\begin{description}

\scalarinteger{maxit} 
of iterations the algorithm is allowed to take before being stopped.  The default value is {\tt 100}.  This is not accessed if {\tt nlls\_iterate} is used.

\scalarinteger{model}
that specifies the model, $m_k(\cdot)$, used.  Possible values are
\begin{description}
  \item{\tt 1} Gauss-Newton (no Hessian).
  \item{\tt 2} (Quasi-)Newton (uses exact Hessian if {\tt exact\_second\_derivatives} is true, otherwise builds an approximation to the Hessian).
%  \item{\tt 3} Barely second-order (Hessian matrix approximated by the identity).
  \item{\tt 3} Hybrid method (mixture of Gauss-Newton/(Quasi-)Newton, as determined by the package).
  \item{\tt 4} Tensor-Newton method.
\end{description}
The default is {\tt model = 3}.

\scalarinteger{type\_of\_method}
that specifies the type of globalization method used.  Possible values are
\begin{description}
  \item{\tt 1} Trust-region method.
  \item{\tt 2} Regularization.
\end{description}
The default is {\tt type\_of\_method = 1}.

\scalarinteger{nlls\_method}
that specifies the method used to solve (or approximate the solution to) the trust-region sub problem.  Possible values are
\begin{description}
  \item{\tt 1} Powell's dogleg method (approximates the solution).
  \item{\tt 2} The Adachi-Iwata-Nakatsukasa-Takeda (AINT) method.
  \item{\tt 3} The More-Sorensen method.
  \item{\tt 4} {\sc Galahad}'s {\sc DTRS} method if {\tt type\_of\_method=1}, or {\sc Galahad}'s {\sc DRQS} method if {\tt type\_of\_method=2}.
\end{description}
If {\tt type\_of\_method = 4}, then only {\tt nlls\_method = 4} is permitted.
The default is {\tt nlls\_method = 4}.

\scalarreal{stop\_g\_absolute}
that specifies the absolute tolerance used in the convergence test on \(\|{\iter{\vJ}}^T\vr(\iter{\vx}))\|/\|\vr(\iter{\vx})\|\).
The default value is {\tt stop\_g\_absolute = 1e{-5}}.

\scalarreal{stop\_g\_relative}
that specifies the relative tolerance used in the convergence test on \(\|{\iter{\vJ}}^T\vr(\iter{\vx})\|/\|\vr(\iter{\vx})\|\).
The default value is {\tt stop\_g\_relative = 1e{-8}}.

\scalarreal{stop\_f\_absolute}
that specifies the absolute tolerance used in the convergence test on \(\|\vr(\iter{\vx})\|\).
The default value is {\tt stop\_f\_absolute = 1e{-8}}.

\scalarreal{stop\_f\_relative}
that specifies the relative tolerance used in the convergence test on \(\|\vr(\iter{\vx})\|\).
The default value is {\tt stop\_f\_relative = 1e{-8}}.

\scalarreal{stop\_s}
that specifies the tolerance used in the convergence test on \(\|\iter{\vs}\|\).
The default value is {\tt stop\_s = eps}.

\scalarinteger{relative\_tr\_radius}
that specifies whether the initial trust region radius should be scaled.
The default value is {\tt relative\_tr\_radius = 0}.

\scalarreal{initial\_radius\_scale}
that specifies the scaling parameter for the initial trust region radius, which is only used if {\tt relative\_tr\_radius = 1}.
The default value is {\tt initial\_radius\_scale = 1.0}.

\scalarreal{initial\_radius}
that specifies the initial trust-region radius, $\Delta$.
The default value is {\tt initial\_radius = 100.0}.

\scalarreal{regularization\_weight}
that specifies the regularization weight, $\sigma$, used in the least-squares problem (\ref{eq:nlls_problem}).  
The default value is {\tt regularization\_weight = 0.0}.

\scalarreal{regularization\_power}
that specifies the regularization weight, $p$, used in the least-squares problem (\ref{eq:nlls_problem}).  
The default value is {\tt regularization\_power = 0.0}.


\scalarreal{maximum\_radius}
that specifies the maximum size permitted for the trust-region radius.
The default value is {\tt maximum\_radius = 1e8}.

\scalarreal{eta\_successful}
that specifies the smallest value of $\rho$ such that the step is accepted.
The default value is {\tt eta\_successful = 1e-8}.

% success_but_reduce is also available, but not documented

\scalarreal{eta\_very\_successful}
that specifies the value of $\rho$ after which the trust-region radius is increased.
The default value is {\tt eta\_very\_successful = 0.9}.

\scalarreal{eta\_too\_successful}
that specifies that value of $\rho$ after which the step is accepted, but keep the trust-region radius unchanged.
The default value is {\tt eta\_too\_successful = 2.0}.

\scalarreal{radius\_increase}
that specifies the factor to increase the trust-region radius by.
The default value is {\tt radius\_increase = 2.0}.

\scalarreal{radius\_reduce}
that specifies the factor to decrease the trust-region radius by.
The default value is {\tt radius\_reduce = 0.5}.

\scalarinteger{tr\_update\_strategy}
that specifies the strategy used to update $\Delta_k$.  Possible values are:
\begin{description}
\item {\tt 1} use the usual step function.
\item {\tt 2} use a the continuous method.
\end{description}
See Section~\ref{sec:step_accept} for more details.
The default value is {\tt 1}.

\scalarreal{hybrid\_switch}
that specifies the value, if {\tt model = 3}, at which second derivatives are used
The default value is {\tt hybrid\_switch = 0.1}.

\scalarlogical{exact\_second\_derivatives}
that, if {\tt true}, signifies that the exact second derivatives are available (and, if {\tt false}, approximates them using a secant method).
The default value is {\tt false}.

\scalarinteger{scale}
that specifies how, if at all, we scale the Jacobian.  We calculate a diagonal scaling matrix, ${\tt D}$, as follows: 
\begin{description}
\item{\tt 0} No scaling is applied ${\tt D} ={\tt I}$.
\item{\tt = 1} Scale using the rows of the Jacobian, \({\tt D}_{i,i} = \|{\tt J}(i,:)\|_2^2\).
\item{\tt = 2} Scale using the rows of the Hessian,  \({\tt D}_{i,i} = \|{\tt H}(i,:)\|_2^2\), where ${\tt H}$ is the (approximation to) the Hessian matrix.
\end{description}
The default value is {\tt scale = 1}

\scalarlogical{scale\_trim\_max}
that specifies whether or not to trim large values of the scaling matrix, ${\tt D}$.
If {\tt true}, ${\tt D}_{i,i} \leftarrow min({\tt D}_{i,i}, {\tt scale\_max})$.
The default value is {\tt true}.

\scalarreal{scale\_max}
that specifies the maximum value allowed if {\tt scale\_trim\_max = true}.
The default value is {\tt 1e11}.

\scalarlogical{scale\_trim\_min}
that specifies whether or not to trim small values of the scaling matrix, ${\tt D}$.
If {\tt true}, ${\tt D}_{i,i} \leftarrow max({\tt D}_{i,i}, {\tt scale\_max})$.
The default value is {\tt true}.

\scalarreal{scale\_min}
that specifies the minimum value allowed if {\tt scale\_trim\_max = true}.
The default value is {\tt 1e-11}.

\scalarlogical{scale\_require\_increase}
that specifies whether or not to require ${\tt D}_{i,i}$ to increase before updating it.
The default value is {\tt false}.

\scalarlogical{calculate\_svd\_J}
that specifies whether or not to calculate the singular value decomposition of ${\tt J}$
at each iteration.  
The default value is {\tt false}.

% \scalarlogical{setup\_workspaces}
% that specifies whether or not to setup the workspaces.  If {\tt setup\_workspaces = false}, then the user must pass a pre-assigned scalar of type {\tt NLLS\_workspace} (generated with the appropriate options) to {\tt nlls\_iterate}, which must come either from a previous call to {\tt nlls\_iterate}.

% \scalarlogical{remove\_workspaces}

\scalarinteger{more\_sorensen\_maxits}
that, if {\tt nlls\_method = 3}, specifies the maximum number of iterations allowed in the More-Sorensen method.
The default value is {\tt more\_sorensen\_maxits = 500}.

\scalarreal{more\_sorensen\_shift}
that, if {\tt nlls\_method = 3}, specifies the shift to be used in the More-Sorensen method.
The default value is {\tt more\_sorensen\_shift = 1e-13}.

\scalarreal{more\_sorensen\_tiny}
that, if {\tt nlls\_method = 3}, specifies the value below which numbers are considered to be essentially zero.
The default value is {\tt more\_sorensen\_tiny = 10.0 * eps}.

\scalarreal{more\_sorensen\_tol}
that, if {\tt nlls\_method = 3}, specifies the tolerance to be used in the More-Sorensen method.
The default value is {\tt more\_sorensen\_tol = 1e-3}.

\scalarreal{hybrid\_tol}
that, if \(\|{\iter{\vJ}}^T \vW \vr(\vx_k) \|_2 < \mathtt{hybrid\_tol} * 0.5 \|\vr(\vx_k)\|_\vW^2\), switches to a \newline(quasi-)Newton method.
The default value is {\tt hybrid\_tol = 2.0}.

\scalarinteger{hybrid\_switch\_its}
that sets how many iterates in a row must the condition in the definition of {\tt hybrid\_tol} hold before a switch.
The default value is {\tt hybrid\_switch\_its = 1}.

\scalarreal{reg\_order}
that, if {\tt nlls\_method = 2}, the order of the regularization used ($p$ in (\ref{eq:reg_subproblem})).  If {\tt reg\_order = 0.0}, then the algorithm
chooses an appropriate value of $p$. The default is {\tt reg\_order = 0.0}.

\scalarinteger{inner\_method} 
that, if {\tt nlls\_method = 4}, specifies the method used to pass in 
the regularization parameter to the inner non-linear least squares solver.  
Possible values are:
\begin{description}
  \item {\tt 1} The current regularization parameter is passed in as a base regularization parameter.
  \item {\tt 2} The regularization term is added to the sum of squares to be solved in the inner problem.
\end{description}
The default is {\tt inner\_method = 2}.

\fortranonly{
\scalarlogical{output\_progress\_vectors}
that, if true, outputs the progress vectors {\tt nlls\_inform\%resvec} and {\tt nlls\_inform\%gradvec} at the end of the routine.
The default value is {\tt false}.
}

\end{description}


%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "nlls_fortran"
%%% End: 
